A multi-scale story of the diffusion of a new technology: the web

### 2024-06-18 paper/submission1/tranos2023.qmd
postpresentation

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)

@@

Geographers were always interested in how new technologies and
innovations diffuse across space and time and, importantly, how such
spatio-temporal processes can be modelled. After all, diffusion together
with invention and innovation are considered the pillars of
technological change [@das2022diffusion]. The seminal contribution of
@hagerstrand1968innovation is illustrative of this early interest.
However, the torch of exploring and modelling such processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space
and time can be attributed to the scarcity of relevant and granular
enough data. As @zook2022mapping highlight, digital activities are
hardly ever captured in official data.

@@

This paper focuses on the diffusion of the web as new technology during
the $1996-2012$ period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
$9$ per cent of UK's households had access to the internet in $1998$
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing was happening exclusively from desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion on a
new technology that is the web. This is key point in the lifecycle of a
new technology. Firstly, during this period new technologies are
expensive, crude and imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this [@tranos2020social]. During this period the performance
of a new technology is the main attraction and not the cost to access
and use it [@wilson2011lessons] There is a broader theoretical
discussion in the literature regarding the motivation behind early
adoption. As summarised by @perkins2005international, on the one hand,
epidemic models highlight the role of interpersonal contacts as a way
for new technologies to diffuse. On the other hand, economic models
underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies because 
of the commonly found lag between investment and economic returns as reflected in 
the Solow paradox [@acemoglu2014return; @brynjolfsson2018artificial].

@@

Importantly, the data used here depicts the *active* engagement with the
web in the UK as it contains all the commercial websites archived by the
Internet Archive [^2] that include a mention to at least one valid UK
postcode in the web text. Websites are identified as commercial ones if
they are part of the .co.uk second level domain (SLD). Also, the act of
of creating a website is understood here as active engagement with the
web vis-à-vis the more passive act of browsing the web or having an
internet connection [@tranosuk]. Previous studies have focused mostly on
more passive notions of engaging with digital technologies such as
internet adoption and internet speeds [e.g. @blank2018local;
@riddlesden2014broadband]. More details about the data and the data
generation process can be found in Section [XX](#sec3).

@@

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise; @grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing]. @grubler1990rise effectively
summarises the three key characteristics of the spatial diffusion
process: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy effect in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the three
mechanisms that the empirical analysis in Section [4]{#sec4} investigates.

@@

**ADD DATA
DESCRIPTION, WHY .CO.UK, FAMILY OF WEB TECHNOLOGIES**

@@

<!-- from our RS paper -->

For instance, we can approximate digital economic activities in a region
by measuring the volume of commercial webpages by considering only
.co.uk websites, which are dedicated to commercial activities (Thelwall,
2000). Such commercial websites are used to exchange information,
support online transactions and share opinions (Blazquez & Domenech,
2018). Although nothing prevents a UK-based company from adopting a
generic TLD such as.com and, indeed, such cases escape our data, we do
not expect that such omissions could affect our results given the
popularity of the .uk ccTLD: UK consumers prefer to visit a .uk website
when they are searching for products or services (Hope, 2017); and
anecdotal evidence indicates that during the first half of 2000, three
.co.uk domains were registered every minute (OECD, 2001). Also, as Table
1 demonstrates, .co.uk was the most popular SLD under the .uk ccTLD in
2000. Hence, we use the total volume of all the archived webpages in a
region as a proxy for the level of digitization. We start the analysis
by focusing on commercial webpages (.co.uk).

@@

# Results {#sec3}

## Old ideas

Hexagon density maps: reflect the spatial structure of Britain. Websites
are associated in places where people live and work.

**TODO** maps at the local authority level per firms and discuss
patterns

Neighbourhood effect: diffusion proceeds outwards from innovation
centers, first "hitting" nearby rather than far-away locations (Grubler
1990)

-   Website density regressions: for OA and LAD over time **TODO** add
    0s Similar pattern

Hierarchy effect: from main centers to secondary ones -- central places

S-shaped diffusion curves

-   S for LADs per firm and OA.

-   Fast and slow LAs map. There is clustering

-   s_uk_firm_11

-   ranks: there is stability and movement

## New structure

@@

-   ~~Website density regressions: for OA and LAD over time~~

@@

When the much more granular scale of OA is adopted, the picture is reversed. The 
most important predictors are the three distance variables to London, the nearest 
city and the nearest retail centre. They stil depict the hierarchical effect, but
proximity to the different population centres is more important than their lagged 
web densities in predicting website diffusion. The neighbouring effect is less 
important at this scale. What is interesting is the almost negligible role of the 
yearly trend and London's website density. While the former probably illustrates 
the large heterogeneity in how web technologies have been adopted at this very fine 
scale, the later highlights that the importance of past web adoption rates in large 
population centres is surpassed by proximity to them and soatial configuration 
at this scale.

>>>>

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)
library(kableExtra)

@@

Geographers were always interested in how new technologies and
innovations diffuse across space and time and, importantly, how such
spatio-temporal processes can be modeled. After all, diffusion together
with invention and innovation are considered the pillars of
technological change [@das2022diffusion]. The seminal contribution of
@hagerstrand1968innovation is illustrative of this early interest.
However, the torch of exploring and modelling such processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space
and time can be attributed to the scarcity of relevant and granular
enough data. As @zook2022mapping highlight, digital activities are
hardly ever captured in official data.

@@

This paper focuses on the diffusion of the web as new technology during
the $1996-2012$ period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
$9$ per cent of UK's households had access to the internet in $1998$
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing was happening exclusively from desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion on a
new technology that is the web as well as its maturity. 
The former is a key point in the lifecycle of a
new technology. Firstly, during this period new technologies are
expensive, crude and imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this [@tranos2020social]; for instance a static website compared 
with a platform like `github`, which enables cooperation between users and 
the creation of new information, meaning, and knowledge 
[@faraj2016special; @barassi2012does]. 
During this period the performance
of a new technology is the main attraction and not the cost to access
and use it [@wilson2011lessons]. There is a broader theoretical
discussion in the literature regarding the motivation behind early
adoption. As summarised by @perkins2005international, on the one hand,
epidemic models highlight the role of interpersonal contacts as a way
for new technologies to diffuse. On the other hand, economic models
underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies because 
of the commonly found lag between investment and economic returns as reflected in 
the Solow paradox [@acemoglu2014return; @brynjolfsson2018artificial].

@@

Importantly, the data used here depicts the *active* engagement with the
web in the UK as it contains all commercial websites that 
(i) are part of the UK's relevant second level domain (SLD, .co.uk), 
(ii) have been archived by the Internet Archive[^2], and 
(iii) include a mention to at least one valid UK
postcode in the web text. 
The act of creating a website is understood here as active engagement with the
web vis-à-vis the more passive act of browsing the web or having an
internet connection [@tranosuk]. Previous studies have focused mostly on
more passive notions of engaging with digital technologies such as
internet adoption and internet speeds [e.g. @blank2018local;
@riddlesden2014broadband]. More details about the data and the data
generation process can be found in Section [XX](#sec3).

@@

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing]. @grubler1990rise effectively
summarises the three key characteristics of the spatial diffusion
process: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy effect in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the three
mechanisms that the empirical analysis in Section [4]{#sec4} investigates.

@@

**ADD FAMILY OF WEB TECHNOLOGIES**

To capture the diffusion of web technologies, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) and the 
Output Areas (OA). The former is an administrative unit and there are 
\\textcolor{red}c. 380\\textcolor{black} such units in the UK. The latter is a census-based
geographical unit, which is very small as there are c. 230,000 of them in the UK.
This methodological choice will allow the mapping of the diffusion 
of web technologies and the assessment of the diffusion mechanisms 
at these two very different spatial scales.

The population of websites is calculated using data from the 
Internet Archive [^2] and, specifically, the JISC UK Web Domain Dataset [@ukwebarchive]. 
The Internet Archive is one of the most complete and oldest archive of webpages 
in the world operating since 1996 [@ainsworth2011much; @holzmann2016dawn].
It is a web crawler, which discovers webpages by following the hyperlinks
of every webpage its archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk country code top-level domain (ccTLD), this 
paper focuses on its commercial subset, the .co.uk SLD. 
This choice decrease the heterogeneity of the web data as 
such commercial websites have rather specific aims: they are used to diffuse information,
support online transactions and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as.com and such cases escape 
the data used here, such omissions should not affect our results given the
popularity of the .uk ccTLD [@tranosuk]: UK consumers prefer to visit a .uk website
when they are searching for products or services [@hope]; and
anecdotal evidence indicates that during the first half of 2000, three
.co.uk domains were registered every minute [@oecd_coms]. 
Importantly, previous studies illustrated that .co.uk is the most popular UK SLD
[@tranosuk].

The text from these webpages was scanned using regular expressions (regex) to
identify strings of text which resemble UK postcodes and one fifth of them included
a mention to a postcode [@BL2013geo]. This information allows the geolocation of the 
data and the creation of the LAD and OA counts.

Before that, the data 
cleaning process included a data aggregation step, through which
the webpages were aggregated to the relevant websites. 
Based on the following example, all three webpages listed below are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)) and at the end 
only websites and not the nested webpages were considered as otherwise the metric 
would have been biased towards large websites.

  - [http://www.website.co.uk/webpage_a](http://www.website.co.uk/webpage_a) B152TT

  - [http://www.website.co.uk/webpage_b](http://www.website.co.uk/webpage_b) BS81TH

  - [http://www.website.co.uk/webpage_c](http://www.website.co.uk/webpage_c) B152TT

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, ([http://www.website.co.uk](http://www.website.co.uk))
includes two unique postcodes: B152TT and BS81SS. 
The distribution of postcodes for 2000 is presented in Table \ref{f2000}, which 
clearly illustrates the wide range. On one hand, we
have websites anchored to a unique location (72% of all
the reconstructed websites), which may represent
a small company with a single trading location. On the other hand,
we have websites with thousands of different postcodes.
Considering the timeperiod of the analysis, such cases can represent directories
which used to be popular in the pre-search engines early times of internet [@tranosuk].[^3]

The analysis presented here is based on two subsets of these data.
Firstly, the analysis is based on websites which only contain only one unique postcode
and, therefore, the geolocation process might suffer from less noise. 
As a robustness check, the analysis is replicated the analysis for an extended subset 
of websites, which include up to ten unique postcodes. These websites are geolocated 
by equally attaching them to multiple locations. This extended sample includes 
94 percent of all the archived websites in 2000.

[^3]: See Figure A1 in **Appendix A in the supplemental data online for examples**.

```{r eval=T, echo=FALSE, results='asis'}

# load 2000 co.uk
# path.2000 <- "C:/Users/nw19521/OneDrive - University of Bristol/projects/archive/nuts/all2000couk.csv"
# path.2000 <- "/Users/nw19521/Library/CloudStorage/OneDrive-UniversityofBristol/projects/archive/nuts/all2000couk.csv"
path.2000 <- paste0(path, "/data/temp/all2000couk.csv")

all2000.duplicates <- data.table::fread(path.2000) #
# dim(all2000.duplicates) # 3336162
# it includes duplicated URLs: if one webpages includes multiple postcodes
# then it appears multiple times. This is ok for the nuts aggregation, but
# not for the frequencies
# This is only .co.uk

# one line for every host
all2000 <- unique(all2000.duplicates, by = c("host")) # 57897

# unique postcodes per website f table
f.websites.pc <- DescTools::Freq(all2000$V1, breaks = c(0, 1,2, 10,100,1000,10000,100000), ord = "desc")
f.websites.pc$level <- factor(f.websites.pc$level, levels = c("[0,1]","(1,2]", "(2,10]", "(10,100]",
                                                              "(100,1e+03]", "(1e+03,1e+04]",
                                                              "(1e+04,1e+05]"))
levels(f.websites.pc$level) <- c("(0,1]","(1,2]", "(2,10]", "(10,100]", "(100,1000]", "(1000,10000]", "(10000,100000]")

f.websites.pc <- f.websites.pc %>% 
  dplyr::select(-cumfreq) %>% 
  rename(Postcodes = level,
         F = freq,
         'F (%)' = perc,
         'Cummulative F' = cumperc)

kable(f.websites.pc,
      format = "latex",
      digits = 3,
      booktabs = T,
      format.args = list(big.mark = ","),
      caption = "Number of unique postcodes per .co.uk website, 2000.\\label{f2000}") %>%
  kableExtra::footnote(general = "Tranos et al. 2021",
                       general_title = "Source: ",
                       footnote_as_chunk = T)  
   # kable_classic(full_width = F, html_font = "Cambria")
```

To create the web density metric, the yearly website counts at the LAD level are
standardised by the number of firms in LAD to avoid biases associated with LADs 
hosting a large number of firms. Given that there is no such statistic for the OA,
the actual OA level counts are used. However, because of the the rather
consistent spatial definition of OA (they host 40-250 households),[^4] 
website counts in OA are interpreted as a density metric too.

[^4]: [https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies](https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies)

This website metric...

@@

# Results {#sec4}

@@

@@

When the much more granular scale of OA is adopted, the picture is reversed. The 
most important predictors are the three distance variables to London, the nearest 
city and the nearest retail centre. They stil depict the hierarchical effect, but
proximity to the different population centres is more important than their lagged 
web densities in predicting website diffusion. The neighbouring effect is less 
important at this scale. What is interesting is the almost negligible role of the 
yearly trend and London's website density. While the former probably illustrates 
the large heterogeneity in how web technologies have been adopted at this very fine 
scale, the later highlights that the importance of past web adoption rates in large 
population centres is surpassed by proximity to them and spatial configuration 
at this scale.
